---
title: Rock Breaks Scissors
author: Michael Foley
date: '2023-11-06'
slug: []
categories:
  - Tutorial
tags: []
description: ''
topics: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(tidyverse)
library(scales)
library(glue)
library(tidymodels)
```

William Poundstone's *Rock Breaks Scissors* ([Amazon](https://www.amazon.com/Rock-Breaks-Scissors-Outguessing-Outwitting/dp/0316228052)) shows how people are often predictable in their efforts to be random. For instance, from the title, a good strategy for playing *Rocks, Paper, Scissors* is to start with *paper* because your opponent is likely to lead with the aggressive *rock*. The insights are neat throughout, but I'm especially delighted with the simple applications of probability. Let's play with a few from the chapter on Chapanis's random number experiment

Chapanis was a noted industrial designer who worked at Bell Labs in the 1950s. His experiment asked students to write a sequence of 2,520 random digits. Chapanis analyzed how random the sequences actually were. I conducted a mini experiment on myself, writing down 300 random numbers before succumbing to tedium.

```{r}
chapanis <- str_c(
  "19654378094657264109869482380864380216091452609914526179532427948267049319418",
  "648091694823753689014549318621563892675943217458272872395417642800142238774519",
  "632514589321792045916134219574162598769412430289014826723940183921825732087428",
  "3109924132384210891423614279413976455277149716169314281753649812435"
)
```

If the sequence is random, the digit frequencies should have a binomial distribution with *p* = 1/10. You expect noise, but this is too much noise.
 
```{r}
tibble(num = 0:9, n = str_count(chapanis, as.character(c(0:9)))) %>%
  mutate(p = n / sum(n)) %>%
  ggplot(aes(x = num, y = p)) + 
  geom_col() +
  geom_text(aes(label = n), nudge_y = -.01, color = "white") +
  scale_x_continuous(breaks = 0:9) +
  scale_y_continuous(labels = percent_format()) +
  geom_hline(yintercept = .10, linetype = 2)
```

Look at digit 0. I wrote it only 18 times (6%). The expected value is 30 (10%). Astonishingly, the least chosen digit by just about everyone in Chapanis's study was 0 - even the way I was non-random was non-random! The probability of writing only 18 zeros in a sequence of 300 digits is <1%.

```{r}
pbinom(q = 18, size = 300, prob = 1/10)
```

If I really chose digits randomly, 95% of experiments like this would yield between 20 and 41 zeros, so I'm on the tail, but to far down it.

```{r}
qbinom(p = c(.025, .975), size = 300, prob = 1/10)
```

It gets better. Chapanis noticed predictable sequences of numbers too. Participants were unlikely to repeat numbers back to back presumably because that feels non-random. On the other hand, descending sequences like "32" and "21" were over-used. Was that true for me?

```{r}
# Convert my 300 digit string into a list of 299 2-digit strings.
my_pairs <- map_chr(1:(str_length(chapanis)-1), ~str_sub(chapanis, ., . + 1)) 

# Frequencies of 2-digit combinations, including combinations I didn't use.
my_freq <- tibble(
  digit_pairs = str_pad(as.character(0:99), 2, "left", "0"),
  freq = map_dbl(digit_pairs, ~sum(my_pairs == .))
  )
```

There were 13 pairs of numbers I did not write at all. I didn't write many zeros, so that accounts for five of the 13, but sure enough, five repeated digits are in the list.

```{r}
my_freq %>% filter(freq == 0) %>% pull(digit_pairs)
```

Only 7 of my 13 most frequently written pairs were descending.

```{r}
my_freq %>% 
  mutate(
    d1 = str_sub(digit_pairs, 1, 1),
    d2 = str_sub(digit_pairs, 2, 2),
    comp = case_when(d1 > d2 ~ "Descending", d1 < d2 ~ "Ascending", TRUE ~ "Repeating")
  ) %>%
  summarize(.by = comp, freq = sum(freq)) %>%
  mutate(pct = freq / sum(freq))
```

The distribution differs from the expected 10% repeats, 45% ascending, 45% descending.

```{r}
chisq.test(x = c(7, 142, 150), p = c(.10, .45, .45))
```

If the digit sequence was random, it should not be possible to predict the next digit in the sequence with more than 1/10 accuracy. However, Chapanis could guess then next digit 17% of the time. Using the prior two digits improved his accuracy to 28%. For me, the prior digit predicted the next digit 22% of the time, and the prior two digits predicted the next digit 31% of the time.

```{r}
dat <- tibble(y = str_split_1(chapanis, ""), l1 = lag(y, 1), l2 = lag(y, 2)) %>%
  mutate(across(everything(), factor))

multinom_reg() %>%
  set_engine("nnet") %>%
  fit(y ~ l1, data = dat) %>%
  augment(new_data = dat) %>%
  summarize(M = mean(.pred_class == y, na.rm = TRUE))

multinom_reg() %>%
  set_engine("nnet") %>%
  fit(y ~ l1 + l2, data = dat) %>%
  augment(new_data = dat) %>%
  summarize(M = mean(.pred_class == y, na.rm = TRUE))
```


 
